---
title: "Style"
author: "Luke Zappia"
date: "`r Sys.Date()`"
output:
    bookdown::word_document2:
        reference_docx: style.docx
bibliography: style_refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

**Lorem ipsum** dolor *sit amet*. 

- Single asterisks italicize text *like this*. 
- Double asterisks embolden text **like this**.

Start a new paragraph with a blank line separating paragraphs.

- This will start an unordered list environment, and this will be the first item.
- This will be a second item.
- A third item.
    - Four spaces and a dash create a sublist and this item in it.
- The fourth item.
    
1. This starts a numerical list.
2. This is no. 2 in the numerical list.
    
# This Starts A New Section
## This is a Subsection
### This is a Subsubsection
#### This starts a Paragraph Block.

> This will create a block quote, if you want one.

Want a table? This will create one.

Table Header  | Second Header
------------- | -------------
Table Cell    | Cell 2
Cell 3        | Cell 4 

Note that the separators *do not* have to be aligned.

Want an image? This will do it.

```{r example-figure, fig.cap = "An example figure"}
par(mar = c(4, 4, 0.1, 0.1))
plot(pressure, pch = 19, type = "b")
```

A normal paragraph.

And here is a cross reference to Figure \@ref(fig:example-figure).

Almost forgot about creating a footnote.[^1] This will do it again.[^2]

[^1]: The first footnote
[^2]: The second footnote

Want to cite something? 

- Find your biblatexkey in your bib file.
- Put an @ before it, like @R-base, or whatever it is.
- @R-base creates an in-text citation (e.g. Smith (1984) says...)
- [@R-base] creates a parenthetical citation (Smith, 1984)

That'll also automatically create a reference list at the end of the document.

[In-text link to Google](http://google.com) as well.
